---
title: "Prompt design"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prompt design}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = elmer:::openai_key_exists() && elmer:::anthropic_key_exists(),
  cache = TRUE
)
options(elmer_seed = 1337)

# Manually ratchet claude variability way down to hopefully make generated
# code better match my prose.
chat_claude <- function(...) {
  elmer::chat_claude(..., api_args = list(temperature = 0))
}
```

In this vignette, you'll learn the basics of writing an LLM prompt, i.e. the text that you send to an LLM asking it to do a job for you. If you've never written a prompt before, a good to way to think about it is as writing a set of instructions for a technically skilled but busy human. You'll need to clearly and concisely state what you want, resolve any potential ambiguities that are likely to arise, and provide a few examples. Don't expect to write the perfect prompt on your first attempt. You'll need to iterate a few times, but in my experience, this iteration is very worthwhile because it forces you to clarify your understanding of the problem.

As well as the general advice in this vignette, it's also a good idea to read the specific advice for the model that you're using. Here are some pointers to the prompt engineering guides for a few popular models:

* [Claude](https://docs.anthropic.com/en/docs/build-with-claude/prompt-engineering/overview)
* [OpenAI](https://platform.openai.com/docs/guides/prompt-engineering)
* [Gemini](https://ai.google.dev/gemini-api/docs/prompting-intro)

If you have a claude account, you can use its <https://docs.anthropic.com/en/docs/build-with-claude/prompt-engineering/prompt-generator>. This prompt generator has been specifically tailored for Claude, but I suspect it will help many other LLMs, or at least give you some ideas as to what else you might want to include in your prompt.

```{r setup}
#| cache: false
library(elmer)
```

## Logistics

There are a few logistics to discuss first. It's highly likely that you will end up writing long, possibly multi-page prompts, so we want to ensure that you're set up for success. So we recommend that you put them in their own file, and write them using markdown. LLMs, like humans, appear to find markdown to be quite readable and markdown allows you to (e.g.) use headers to divide up the prompt into esection, and other tools like itemised lists to enumerate multiple options. You can see some a few examples of this style of prompt at the following urls:

* <https://github.com/posit-dev/shiny-assistant/blob/main/shinyapp/app_prompt_python.md>
* <https://github.com/jcheng5/py-sidebot/blob/main/prompt.md>
* <https://github.com/simonpcouch/pal/tree/main/inst/prompts>
* <https://github.com/cpsievert/aidea/blob/main/inst/app/prompt.md>

If you only have one prompt in your project, call it `prompt.md`. If you have multiple prompts, give them informative names like `prompt-extract-metadata.md` or `prompt-summarize-text.md`. If you're writing a package, put your prompt(s) in `inst/prompts`, otherwise it's fine to put them in the root directory of your project.

Your prompts are going to change over time, so we highly recommend using them with git. That will ensure that you can easily see what has changed, and if you accidentally make a mistake you can easily roll back to a known good verison.

If your prompt includes dynamic data, `elmer::interpolate_file()` to interpolate it into your prompt. `interpolate_file()` works like [glue](https://glue.tidyverse.org), but uses `{{ }}` instead of `{ }` to make it easier to work with JSON.

As you iterate on the prompt, it's a good idea to build up a small set of challenging examples that you can regularly re-check with your latest version of the prompt. Currently you'll need to do this by hand, but we hope to eventually also provide tools that help you do this a little more formally.

Unforunatey you won't see these logistics in action in this vignette, since we're keeping the prompts short and inline to make it easier for you to grok what's going on.

## Code generation

Let's explore prompt design for a simple code generation task:

```{r}
#| cache: false
question <- "
  How can I compute the mean and median of variables a, b, c, and so on,
  all the way up to z, grouped by age and sex.
"
```

I'll use `chat_claude()` for this problem because in our experience it does the best job of generating code.

### Basic flavour

When I don't provide a system prompt, I sometimes get answers in a different languages or different styles of R code:

```{r}
#| label: code-basic
chat <- chat_claude()
chat$chat(question)
```

I can ensure that I always get R code in a given by providing a system prompt:

```{r}
#| label: code-r
chat <- chat_claude(system_prompt = "
  You are an expert R programmer who prefers the tidyverse.
")
chat$chat(question)
```

Note that I'm using both a system prompt (which defines the general behaviour) and a user prompt (which askes the specific question). You could put all of the content in the user prompt and get similar results, but I think it's helpful to use both to cleanly divide the general framing of the response from the specific questions that you want to ask.

Since I'm mostly interested in the code, I ask it to drop the explanation and the sample data:

```{r}
#| label: code-r-minimal
chat <- chat_claude(system_prompt = "
  You are an expert R programmer who prefers the tidyverse.
  Just give me the code. I don't want any explanation or sample data.
")
chat$chat(question)
```

If you want a different style of R code, you can of course ask for it:

```{r}
#| label: code-styles
chat <- chat_claude(system_prompt = "
  You are an expert R programmer who prefers data.table.
  Just give me the code. I don't want any explanation or sample data.
")
chat$chat(question)

chat <- chat_claude(system_prompt = "
  You are an expert R programmer who prefers base R.
  Just give me the code. I don't want any explanation or sample data.
")
chat$chat(question)
```

### Be explicit

If there's something about the output that you don't like, you can try being more explicit about it. For example, the code isn't styled quite how I like, so I provide more details about what I do want:

```{r}
#| label: code-explicit
chat <- chat_claude(system_prompt = "
  You are an expert R programmer who prefers the tidyverse.
  Just give me the code. I don't want any explanation or sample data.

  Follow the tidyverse style guide:
  * Spread long function calls across multiple lines.
  * Where needed, always indent function calls with two spaces.
  * Only name arguments that are less commonly used.
  * Always use double quotes for strings.
  * Use the base pipe, `|>`, not the magrittr pipe `%>%`.
")
chat$chat(question)
```

This still doesn't yield exactly the code that I'd write, but it's prety close.

You could provide a different prompt if you were looking for more explanation of the code:

```{r}
#| label: code-teacher
chat <- chat_claude(system_prompt = "
  You are an expert R teacher.
  I am a new R user who wants to improve my programming skills.
  Help me understand the code you produce by explaining each function call with
  a brief comment. For more complicated calls, add documentation to each
  argument. Just give me the code. I don't want any explanation or sample data.
")
chat$chat(question)
```

### Teach it about new features

You can imagine LLMs as being a sort of an average of the internet at a given point in time. That means they will provide popular answers, which will tend to reflect older coding styles (either because the new features aren't in their index, or the older features are so much more popular). So if you want your code to use specific features that are relatively recent, you might need to provide the examples yourself:

```{r}
#| label: code-new-feature
chat <- chat_claude(system_prompt = "
  You are an expert R programmer.
  Just give me the code; no explanation in text.
  Use the `.by` argument rather than `group_by()`.
  dplyr 1.1.0 introduced per-operation grouping with the `.by` argument.
  e.g., instead of:

  transactions |>
    group_by(company, year) |>
    mutate(total = sum(revenue))

  write this:
  transactions |>
    mutate(
      total = sum(revenue),
      .by = c(company, year)
    )
")
chat$chat(question)
```

## Structured data

Providing a rich set of examples is a great way to encourage the output to produce exactly what you want. This is also known as **multi-shot prompting**. Here we'll work through a prompt that I designed to extract structured data from recipes, but the same ideas apply in many other situations.

### Getting started

My overall goal is to turn a list of ingredients, like the following, into a nicely structured JSON that I can then analyse in R (e.g. to compute the total weight, scale the recipe up or down, or to convert the units from volumes to weights).

```{r}
#| cache: false
ingredients <- "
  ¾ cup (150g) dark brown sugar
  2 large eggs
  ¾ cup (165g) sour cream
  ½ cup (113g) unsalted butter, melted
  1 teaspoon vanilla extract
  ¾ teaspoon kosher salt
  ⅓ cup (80ml) neutral oil
  1½ cups (190g) all-purpose flour
  150g plus 1½ teaspoons sugar
"
```

(This isn't the ingredient list for a real recipe but it includes a sampling of styles that I encountered in my project.)

If you don't have strong feelings about what the data structure should look like, you can start with a very loose prompt and see what you get back. I find this a useful pattern for underspecified problems where a big part of the problem is just defining precisely what problem you want to solve. Seeing the LLMs attempt at coming up with a data structure gives me something to immediately react to, rather than having to start from a blank page.

```{r}
#| label: data-loose
instruct_json <- "
  You're an expert baker who also loves JSON. I am going to give you a list of
  ingredients and your job is to return nicely structured JSON. Just return the
  JSON and no other commentary.
"

chat <- chat_openai(instruct_json)
chat$chat(ingredients)
```

(I don't know if the colour text, "You're an expert baker who also loves JSON", does anything, but I like to think this helps the LLM get into the right mindset of a very nerdy baker.)

### Provide examples

This isn't a bad start, but I prefer to cook with weight, so I only want to see volumes if weight isn't available. So I provide a couple of examples of what I'm looking for. I was pleasantly suprised that I can provide the input and output examples in such a loose format.

```{r}
#| label: data-examples
instruct_weight <- r"(
  Here are some examples of the sort of output I'm looking for:

  ¾ cup (150g) dark brown sugar
  {"name": "dark brown sugar", "quantity": 150, "unit": "g"}

  ⅓ cup (80ml) neutral oil
  {"name": "neutral oil", "quantity": 80, "unit": "ml"}

  2 t ground cinnamon
  {"name": "ground cinnamon", "quantity": 2, "unit": "teaspoon"}
)"

chat <- chat_openai(c(instruct_json, instruct_weight))
chat$chat(ingredients)
```

Just providing the examples seems to work remarkably well. But I found it useful to also include description of what the examples are trying to accomplish. I'm not sure if this helps the LLM or not, but it certainly makes it easier for me to understand the organisation and check that I've covered the key pieces that I'm interested in.

```{r}
#| cacahe: false
instruct_weight <- r"(
  * If an ingredient has both weight and volume, extract only the weight:

  ¾ cup (150g) dark brown sugar
  [
    {"name": "dark brown sugar", "quantity": 150, "unit": "g"}
  ]

* If an ingredient only lists a volume, extract that.

  2 t ground cinnamon
  ⅓ cup (80ml) neutral oil
  [
    {"name": "ground cinnamon", "quantity": 2, "unit": "teaspoon"},
    {"name": "neutral oil", "quantity": 80, "unit": "ml"}
  ]
)"
```

This structure also allows me to give the LLMs a hint about how I want multiple ingredients to be stored, i.e. as an JSON array.

I then just iterated on this task, looking at the results from different recipes to get a sense of what the LLM was getting wrong. Much of this felt like I waws iterating on my understanding of the problem as I didn't start by knowing exactly how I wanted the data. For example, when I started out I didn't really think about all the various ways that ingredients are specified. For later analysis, I always want quantities to be number, even if they were originally fractions, or the if the units aren't precise (like a pinch). It also forced me to realise that some ingredients are unitless.

```{r}
#| cache: false
instruct_unit <- r"(
* If the unit uses a fraction, convert it to a decimal.

  ⅓ cup sugar
  ½ teaspoon salt
  [
    {"name": "dark brown sugar", "quantity": 0.33, "unit": "cup"},
    {"name": "salt", "quantity": 0.5, "unit": "teaspoon"}
  ]

* Quantities are always numbers

  pinch of kosher salt
  [
    {"name": "kosher salt", "quantity": 1, "unit": "pinch"}
  ]

* Some ingredients don't have a unit.
  2 eggs
  1 lime
  1 apple
  [
    {"name": "egg", "quantity": 2},
    {"name": "lime", "quantity": 1},
    {"name", "apple", "quantity": 1}
  ]
)"
```

You might want to take a look at the [full prompt](https://gist.github.com/hadley/7688b4dd1e5e97b800c6d7d79e437b48) to see what I ended up with.

### Structured data

Now that I've iterated to get a data structure that I like, it seems useful to formalise it and tell the LLM exactly what I'm looking for using structured data. This guarantees that the LLM will only return JSON, the JSON will have the fields that you expect, and then elmer will automatically convert it into an R data structure for you.

```{r}
#| label: data-structured
type_ingredient <- type_object(
  name = type_string("Ingredient name"),
  quantity = type_number(),
  unit = type_string("Unit of measurement", required = FALSE)
)

type_ingredients <- type_array(items = type_ingredient)

chat <- chat_openai(c(instruct_json, instruct_weight))
data <- chat$extract_data(ingredients, spec = type_object(ingredients = type_ingredients))
do.call(rbind, lapply(data$ingredients, as.data.frame))
```

### Capturing raw input

One thing that I'd do next time would also be to include the raw ingredient name in the output. This doesn't make much difference here, in this simple example, but it makes it much easier to align the input and the output and start to develop automated measures of how well my prompt is doing.

```{r}
#| cache: false
instruct_weight_input <- r"(
  * If an ingredient has both weight and volume, extract only the weight:

    ¾ cup (150g) dark brown sugar
    [
      {"name": "dark brown sugar", "quantity": 150, "unit": "g", "input": "¾ cup (150g) dark brown sugar"}
    ]

  * If an ingredient only lists a volume, extract that.

    2 t ground cinnamon
    ⅓ cup (80ml) neutral oil
    [
      {"name": "ground cinnamon", "quantity": 2, "unit": "teaspoon", "input": "2 t ground cinnamon"},
      {"name": "neutral oil", "quantity": 80, "unit": "ml", "input": "⅓ cup (80ml) neutral oil"}
    ]
)"
```

I think this is particularly important if you're working with even less structured text. For example, imagine you had this text:

```{r}
#| cache: false
recipe <- r"(
  In a large bowl, cream together one cup of softened unsalted butter and a
  quarter cup of white sugar until smooth. Beat in an egg and 1 teaspoon of
  vanilla extract. Gradually stir in 2 cups of all-purpose flour until the
  dough forms. Finally, fold in 1 cup of semisweet chocolate chips. Drop
  spoonfuls of dough onto an ungreased baking sheet and bake at 350°F (175°C)
  for 10-12 minutes, or until the edges are lightly browned. Let the cookies
  cool on the baking sheet for a few minutes before transferring to a wire
  rack to cool completely. Enjoy!
)"
```

Including the input text in the output makes it easier to see if it's doing a good job:

```{r}
#| label: data-unstructured-input
chat <- chat_openai(c(instruct_json, instruct_weight_input))
chat$chat(recipe)
```

When I ran it while writing this vignette, it seems to be working out the weight of the ingredients specified in volume, even though the prompt specifically asks it not to do that. This may suggest I need to broaden my examples.

## Token usage

```{r}
#| label: usage
#| type: asis
#| echo: false
knitr::kable(token_usage())
```
